<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{R/qtl2 input file format}
-->

# R/qtl2 input file format

[R/qtl2](https://kbroman.org/qtl2) (aka qtl2) is a reimplementation of the QTL analysis software
[R/qtl](http://www.rqtl.org), to better handle high-dimensional data
and complex cross designs.

The input data file formats for [R/qtl](http://www.rqtl.org) cannot
handle complex crosses, and so for R/qtl2, we need to define a new
input file format. This document describes the details.

For simple cross types, we can continue to use the file formats for
[R/qtl](http://www.rqtl.org), use `qtl::read.cross()` to read in the
data, and then use a conversion function (`qtl2::convert2cross2()`) to convert
the data into the new format.

For more complex crosses, we need to define a new format. I was
persuaded by [Aaron Wolen](http://aaronwolen.com/)'s idea of a
[&ldquo;tidy&rdquo; format](https://github.com/kbroman/qtl/pull/20)
for [R/qtl](http://www.rqtl.org), with three separate CSV files, one
for phenotypes, one for genotypes, and one for the genetic map.

An another important idea is from [Pjotr Prins](http://thebird.nl/)'s
[qtab format](https://github.com/qtlHD/qtlHD/blob/master/doc/input/qtab.md): 
the inclusion of metadata, such as genotype
encodings, with the primary data. This will simplify the handling of
multiple files and will help to avoid mistakes.

And so the basic idea for the new format is to have a separate file
for each part of the primary data (genotypes, founder genotypes,
genetic map, physical map, phenotypes, covariates, and phenotype
covariates), and then a control file which specifies the names of all
of those files, the genotype encodings and missing value codes, and
things like the name of the sex column within the covariate data (and
the encodings for the sexes) and which chromosome is the X chromosome.

## Reading the data files

Before discussing the boring file specifications, let's consider briefly
how the data are read into R.

A key advantage of the control file scheme is that it greatly
simplifies the function for reading in the data. That function,
`read_cross2()`, has a _single_ argument: the name (with path) of the
control file. So you can read in data like this:

```{r read_cross2, eval=FALSE}
library(qtl2)
grav2 <- read_cross2("~/my_data/grav2.yaml")
```

The large number of files is a bit cumbersome, so we've made it
possible to use a
[zip file](http://en.wikipedia.org/wiki/Zip_(file_format) containing
all of the data files, and to read that zip file directly. There's even a
function for creating the zip file:

```{r create_zip, eval=FALSE}
zip_datafiles("~/my_data/grav2.yaml")
```

The `zip_datafiles()` function will read the control file to identify
all of the relevant data files and then zip them up into a file with
the same name and location, but with the extension `.zip` rather than
`.yaml`.

To read the data back in, we use the same `read_cross2()` function,
providing the name (and path) of the zip file rather than the control
file.

```{r read_zip, eval=FALSE}
grav2 <- read_cross2("~/my_data/grav2.zip")
```

This can even be done with remote files.

```{r read_remote_zip, eval=FALSE}
grav2 <- read_cross2("http://kbroman.org/qtl2/assets/sampledata/grav2/grav2.zip")
```

Of course, the other advantage of the zip file is that it is
_compressed_ and so smaller than the combined set of CSV files.

## Format of the data files

### YAML control file

(Mention here the use of `write_control_file()` for creating these
YAML files.)

### Genotype data

### Founder genotype data

### Phenotype data

### Covariate data

### Phenotype covariates

### Genetic map

### Physical map

### Cross_info file

## Detailed specifications for each cross type


<!-- the following to make it look nicer -->
<link href="http://kbroman.org/qtlcharts/assets/vignettes/vignette.css" rel="stylesheet"></link>
