---
title: "Genotype diagnostics in diversity outbred mice"
output:
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE, warning=FALSE,
                      fig.width=11, fig.height=6)
options(scipen=10, width=110)
```

QTL analysis results are dependent upon the fidelity of the data, and
so an important preliminary activity is the careful investigation of
the quality of both the genotypes and phenotypes. Diagnosing genotype
data problems can be particularly tricky in multi-parent populations,
such as Diversity Outbred (DO) mouse populations. In this document, we
illustrate our basic strategies for genotype diagnostics in DO mice.

We will consider, as our example, a subset of the DO data from [Gatti
et al. (2017)](https://doi.org/10.1101/098657).

The genotype data are for the a set of 291 mice from generations 8 and
11, with genotypes from the
[MegaMUGA array](http://genomics.neogen.com/en/mouse-universal-genotyping-array).
The primary genotype data are in two `FinalReport.txt` files, in the form provided by
GeneSeek, available at <https://doi.org/10.6084/m9.figshare.7359542.v1>.
The data have been converted to
[R/qtl2](http://kbroman.org/qtl2) format following the document
["Preparing Diversity Outbred (DO) mouse data for R/qtl2"](http://kbroman.org/qtl2/pages/prep_do_data.html).
The processed data are available
at <https://github.com/rqtl/qtl2data/tree/master/DO_Svenson291>.

We first load the [R/qtl2](https://github.com/rqtl/qtl2)
package and the data. We'll also load the
[R/broman](https://github.com/kbroman/broman) package for some
utilities and plotting functions, and
[R/qtlcharts](https://kbroman.org/qtlcharts) for interactive
graphs. The package [devtools](https://github.com/r-lib/devtools) is used for `session_info()` at the
very end of the document.

```{r load_data, eval=FALSE}
library(broman)
library(qtl2)
library(qtlcharts)
library(devtools)
file <- paste0("https://raw.githubusercontent.com/rqtl/",
               "qtl2data/master/DO_Svenson291/svenson.zip")
svenson <- read_cross2(file)
```

```{r load_data_really, echo=FALSE}
library(broman)
library(qtl2)
library(qtlcharts)
set.seed(94238567) # set random number seed

url <- paste0("https://raw.githubusercontent.com/rqtl/",
              "qtl2data/master/DO_Svenson291/svenson.zip")
rds_file <- "diag_cache/svenson.rds"
if(!dir.exists("diag_cache")) dir.create("diag_cache")
if(!file.exists(rds_file)) {
    svenson <- read_cross2(url)
    saveRDS(svenson, rds_file)
} else {
    svenson <- readRDS(rds_file)
}
```

I'm going to omit markers without any genotype data and re-code the
genotypes so that the first allele is that which is most common among
the founders.

```
svenson <- drop_nullmarkers(svenson)

for(chr in seq_along(svenson$founder_geno)) {
    fg <- svenson$founder_geno[[chr]]
    g <- svenson$geno[[chr]]
    f1 <- colSums(fg==1)/colSums(fg != 0)

    fg[fg==0] <- NA
    g[g==0] <- NA

    fg[,f1 < 0.5] <- 4 - fg[,f1 < 0.5]
    g[,f1 < 0.5]  <- 4 - g[,f1 < 0.5]

    fg[is.na(fg)] <- 0
    g[is.na(g)] <- 0

    svenson$founder_geno[[chr]] <- fg
    svenson$geno[[chr]] <- g
}
```

We are left with `r add_commas(tot_mar(svenson))` markers
genotyped on `r n_ind(svenson)` mice.


## Missing data per sample

We find it best to first look at the proportion of missing genotypes per
sample. Samples with appreciable missing data (low call rates) are
likely bad, and they'll show up repeatedly as outliers in subsequent
analyses. We can use the R/qtl2 function `n_missing()` to get the
proportion of missing data by sample.

```{r percent_missing}
percent_missing <- n_missing(svenson, "ind", "prop")*100
```

Here's a plot of the percent missing genotypes, by mouse index. The
plot is interactive in that if you hover over a point, you'll view the
individual IDs and the percent missing genotypes for that individual.

```{r iplot_percent_missing, fig.width=8.8, fig.height=4.8}
labels <- paste0(names(percent_missing), " (", round(percent_missing), "%)")
iplot(seq_along(percent_missing), percent_missing, indID=labels,
      chartOpts=list(xlab="Mouse", ylab="Percent missing genotype data",
                     ylim=c(0, 100)))
```

There are `r sum(percent_missing >= 19.97)` mice that are missing >20%
genotypes. Another `r sum(percent_missing > 5 & percent_missing < 19.97)`
mice are missing >5%.

Some of these samples should probably be omitted, but we'll leave them
in for now.


## Sexes


We next seek to verify the sexes of the mice.
One can look at
the proportion of heterozygous genotype calls on the X chromosome, but
we find it most informative to look at the SNP array intensities for
SNPs on the X and Y chromosomes.

When converting the `FinalReport.txt` files to R/qtl2 format, we pulled
out the average intensities for each SNP on the X and Y chromosomes.
We load them into R as follows.

```{r load_XYintensities, eval=FALSE}
url <- paste0("https://raw.githubusercontent.com/rqtl/",
              "qtl2data/master/DO_Svenson291/)
xint <- read_csv_numer(paste0(url, "svenson_chrXint.csv"), transpose=TRUE)
yint <- read_csv_numer(paste0(url, "svenson_chrYint.csv"), transpose=TRUE)
```

```{r load_XYintensities_really, echo=FALSE}
url <- paste0("https://raw.githubusercontent.com/rqtl/",
              "qtl2data/master/DO_Svenson291/")
local_dir <- "diag_cache/"
xfile <- "svenson_chrXint.csv"
yfile <- "svenson_chrYint.csv"
dist_xfile <- paste0(url, xfile)
local_xfile <- paste0(local_dir, xfile)
dist_yfile <- paste0(url, yfile)
local_yfile <- paste0(local_dir, yfile)
if(!file.exists(local_xfile)) download.file(dist_xfile, local_xfile)
if(!file.exists(local_yfile)) download.file(dist_yfile, local_yfile)
xint <- read_csv_numer(local_xfile, transpose=TRUE)
yint <- read_csv_numer(local_yfile, transpose=TRUE)
```

We'll omit data for the `r sum(percent_missing >= 19.97)` samples we've
just omitted.

```{r omit_samples_from_intensities}
xint <- xint[ind_ids(svenson),]
yint <- yint[ind_ids(svenson),]
```

In this project, the sex of the mice was encoded as the first character in the
individual identifiers, so we pull those out, as values `"M"` and
`"F"` for males and females, respectively.

```{r get_sex}
sex <- substr(rownames(xint), 1, 1)
```

Some of the SNPs do not appear to be informative about mouse sex, and
so we first perform t-tests at each SNP, to identify SNPs that show a
large sex difference in intensity.

```{r snps_sex_diff, eval=FALSE}
x_pval <- apply(xint, 2, function(a) t.test(a ~ sex)$p.value)
y_pval <- apply(yint, 2, function(a) t.test(a ~ sex)$p.value)
```

```{r snps_sex_diff_really, echo=FALSE}
file <- "diag_cache/snps_sex_diff.RData"
if(file.exists(file)) {
    load(file)
} else {
<<snps_sex_diff>>
    save(x_pval, y_pval, file=file)
}
```

We then calculate the average intensities on the X and Y chromosomes,
among SNPs that showed a P-value < 0.05 after a Bonferroni correction
for multiple testing.

```{r ave_XYint}
xint_ave <- rowMeans(xint[, x_pval < 0.05/length(x_pval)], na.rm=TRUE)
yint_ave <- rowMeans(yint[, y_pval < 0.05/length(y_pval)], na.rm=TRUE)
```

The following is an interactive scatterplot of the average SNP intensity
on the Y chromosome versus the average SNP intensity on the X
chromosome. Male mice are in purple, and female mice are in green.
If you hover over a point, you'll see the individual ID and the percent
missing genotypes for that individual.

```{r plot_xyint, fig.width=8.8, fig.height=4.8}
point_colors <- as.character( brocolors("web")[c("green", "purple")] )
percent_missing <- n_missing(svenson, summary="proportion")*100
labels <- paste0(names(xint_ave), " (", round(percent_missing), "%)")
iplot(xint_ave, yint_ave, group=sex, indID=labels,
      chartOpts=list(pointcolor=point_colors, pointsize=4,
                     xlab="Average X chr intensity", ylab="Average Y chr intensity"))
```

There's a distinct cluster of male mice in the upper-left (low X
chromosome intensity and high Y chromosome intensity), and a cluster
of female mice in the lower-right (high X chromosome intensity and low
Y chromsome intensity).

There's one male mouse, `r names(xint_ave)[sex=="M" & xint_ave > 0.45]`,
who appears within the female cluster in the lower right, and so is
likely really female.

There's one female mouse, `r names(xint_ave)[sex=="F" & xint_ave < 0.45]`,
in the lower-left, with reduced X chromosome intensity. This may be an
XO female.

Also note that the male samples that are to the left and below the
main cluster (reduced intensities on both the X and Y chromosomes)
are samples with higher rates of missing data. There's also a female
sample with higher-than-normal Y chromosome intensity; this sample
also has a high rate of missing data.

The proportion of heterozygous genotypes on the X chromosome is also
informative, but less so. In the following scatterplot, we show the proportion of
hets vs the average intensity for the X chromosome SNPs. (Note that in
calculating the proportion of heterozygous genotypes for the
individuals, we look at X chromosome genotypes equal to
2 (which corresponds to the heterozygote) relative to not being 0
(which is used to encode missing genotypes). And the genotypes are
arranged with rows being individuals and columns being markers.)

```{r plot_phet_Xchr, fig.width=8.8, fig.height=4.8}
phetX <- rowSums(svenson$geno$X == 2)/rowSums(svenson$geno$X != 0)
iplot(xint_ave, phetX, group=sex, indID=labels,
      chartOpts=list(pointcolor=point_colors, pointsize=4,
                     xlab="Average X chr intensity", ylab="Proportion het on X chr"))
```

The males all show low heterogyzosity on the X chromosome, except for
the one sample (`r names(xint_ave)[sex=="M" & xint_ave > 0.45]`) that
looks to be really female. There are several male samples with
heterozygosity above 5%, but these are again the samples with high
rates of missing genotypes.

The females all show heterozygosity well above 0, except for the one
mouse (`r names(xint_ave)[sex=="F" & xint_ave < 0.45]`) that looks to
be XO.

In summary, we conclude that sample M377 is really for a female mouse,
and sample F386 is likely XO.


## Sample duplicates

We next look for possible sample duplicates by comparing the proportion of
matching SNP genotypes among all pairs of individuals, using the
`compare_geno()` function in
[R/qtl2](https://github.com/rqtl/qtl2). We use `cores=0` to
speed up the calculations using multiple CPU cores in parallel. The
`summary()` function shows pairs that share more than 90% of genotypes.

```{r compare_genotypes}
cg <- compare_geno(svenson, cores=0)
summary(cg)
```

There are two pairs that share almost identical genotypes: M283/M292, and
M377/F409. Note that M377 was identified above, as a male mouse
that looked to really be a female mouse.

Here is a histogram of the proportion of matching genotypes.
The tick marks below the histogram indicate individual pairs.

```{r hist_prop_matching, dev="png", doi=600}
par(mar=c(5.1,0.6,0.6, 0.6))
hist(cg[upper.tri(cg)], breaks=seq(0, 1, length=201),
     main="", yaxt="n", ylab="", xlab="Proportion matching genotypes")
rug(cg[upper.tri(cg)])
```

The figure is made more complicated by the presence of the samples
with high rates of missing data. If we omit the `r sum(percent_missing >= 50)`
samples with > 50% missing genotypes, we get the following figure:

```{r hist_prop_matching_omit3, dev="png", doi=600}
cgsub <- cg[percent_missing < 50, percent_missing < 50]
par(mar=c(5.1,0.6,0.6, 0.6))
hist(cgsub[upper.tri(cgsub)], breaks=seq(0, 1, length=201),
     main="", yaxt="n", ylab="", xlab="Proportion matching genotypes")
rug(cgsub[upper.tri(cgsub)])
```

The majority of pairs share about 50% genotypes, but there
is a group of pairs that are more closely related and share about 70%
genotypes. The two pairs cited above are the only ones with >90%
matching genotypes. There's one other pair with just under 90%
matching genotypes, F326 and F336, but note that F326 is missing
`r round(percent_missing["F326"])`% genotypes.

In summary, we have two clear pairs of sample duplicates: M283/M292, and
M377/F409, with the latter likely being F409 (since both samples
look like a female mouse).

So let's omit sample M377 plus one of M283/M292. (For the latter pair,
we can't be sure which is correct; for later QTL analysis, we might
want to omit both of them, but for our present purposes we can just
omit one of the two at random. Flip of a coin...we'll omit M292.)

```{r omit_sample_duplicates}
svenson <- svenson[c("-M377", "-M292"),]
```

```{r save_svenson_v2, echo=FALSE}
file <- "diag_cache/svenson_v2.rds"
if(!file.exists(file)) {
    saveRDS(svenson, file)
}
```

We are left with `r n_ind(svenson)` individuals, though still including
`r sum(n_missing(svenson, "ind", "prop") > 0.199)` that are missing > 20%
genotypes.


## Bad samples

Our first step in studying these data was to look at the proportion of
missing genotypes per sample. We found `r sum(percent_missing>=19.97)`
samples with >20% missing data. The amount of missing genotypes is the
key indicator of poor sample quality.

Here are the top 19 samples, by the amount of missing data. In
addition to the `r sum(percent_missing>=19.97)` that are missing > 20%
genotypes, there are
`r sum(percent_missing>5 & percent_missing<19.97)` that are missing
around 9% of their genotypes, and
`r sum(percent_missing>3 & percent_missing<8.9)` that are missing
3-5%.

```{r percent_missing_again}
percent_missing <- n_missing(svenson, "ind", "prop")*100
round(sort(percent_missing, decreasing=TRUE)[1:19], 1)
```

### Array intensities

Another useful indicator is
the distribution of intensities on the MegaMUGA arrays. For speed of
access, we store the array intensities in an `.fst` file, with the R
package [fst](http://www.fstpackage.org/), as a matrix of SNPs x
samples, with the two alleles at a SNP being stored in adjacent rows.
See the R script [`geneseek2fst.R`](https://kbroman.org/qtl2/assets/geneseek2fst.R)
for the details of how we converted the
`FinalReport.txt` files from GeneSeek into this format.

We'll read in the full set of intensities and take the sum of the two
allele intensities at each SNP.

```{r read_intensities}
int <- fst::read.fst("diag_cache/intensities.fst")
int <- int[seq(1, nrow(int), by=2),-(1:2)] + int[-seq(1, nrow(int), by=2),-(1:2)]
int <- int[,ind_ids(svenson)]
```

In the following plot, we display the distributions of
array intensities (after a $\log_{10}(x+1)$ transformation).

In the
top panel, the arrays are sorted by the proportion of missing genotype
data for the sample, and the curves
connect various quantiles of the intensities. Hover over the top
panel, and the corresponding histogram is shown below.

```{r iboxplot, fig.height=6.4, fig.width=8.8}
n <- names(sort(percent_missing, decreasing=TRUE))
iboxplot(log10(t(int[,n])+1), orderByMedian=FALSE, chartOpts=list(ylab="log10(SNP intensity + 1)"))
```

The first `r numbers[sum(percent_missing >= 19.97)]` arrays are the ones
that are missing > 20% of the genotypes; they all (well, except for
F326) show a reduced median and long right tail. The next three arrays
are not so extreme, but still show greater variability in array
intensity than the bulk of the arrays, and F326 looks a bit like
these. There are then a set of samples showing a spike of low
intensity values.

For this particular set of arrays, a plot of the 1 %ile vs the 99 %ile
is quite revealing. In the following, the orange points are those with >
20% missing genotypes, the pink points are the samples with 5-20%
missing genotypes, and the blue points are the samples with 2-5%
missing genotypes.

```{r intensity_quantiles, eval=FALSE}
qu <- apply(int, 2, quantile, c(0.01, 0.99), na.rm=TRUE)
group <- (percent_missing >= 19.97) + (percent_missing > 5) + (percent_missing > 2) + 1
labels <- paste0(colnames(qu), " (", round(percent_missing), "%)")
iplot(qu[1,], qu[2,], indID=labels, group=group,
      chartOpts=list(xlab="1 %ile of array intensities",
                     ylab="99 %ile of array intensities",
                     pointcolor=c("#ccc", "slateblue", "Orchid", "#ff851b")))
```

```{r intensity_quantiles_really, echo=FALSE, fig.width=8.8, fig.height=4.8}
file <- "diag_cache/array_qu.rds"
if(file.exists(file)) {
    qu <- readRDS(file)
} else {
    qu <- apply(int, 2, quantile, c(0.01, 0.99), na.rm=TRUE)
    saveRDS(qu, file)
}
group <- (percent_missing >= 19.97) + (percent_missing > 5) + (percent_missing > 2) + 1
labels <- paste0(colnames(qu), " (", round(percent_missing), "%)")
iplot(qu[1,], qu[2,], indID=labels, group=group,
      chartOpts=list(xlab="1 %ile of array intensities",
                     ylab="99 %ile of array intensities",
                     pointcolor=c("#ccc", "slateblue", "Orchid", "#ff851b")))
```

The samples with >20% missing data (in orange) have somewhat reduced 1st percentile and
elevated 90th percentile, except for one orange point (F326) that
is close to the bulk of the samples. The samples with 5-9% missing data (in pink)
are similar but less extreme. Many of the samples with 2-5% missing
data show decreased 1st percentile, but many look fit in with the bulk
of the samples.

We're tempted to omit the `r sum(percent_missing > 5)` samples with >5%
missing data, but we'll continue to leave them in, for now.


### Genotype frequencies

It can be useful to look at the genotype frequencies for the raw
SNP data. We'll focus on the autosomal
markers.

We first pull out the SNP genotypes. They've been coded according
to the allele frequencies in the founder strains, with 1 and 3
corresponding to the homozygotes for the major and minor alleles,
respectively; let's call them AA and BB (that is, A is the major allele).
We also pull out the founder genotypes and count the number of
founders with the minor allele. In studying the genotype frequencies,
we'll omit markers where any of the founders' genotypes are missing
(coded as 0).

```{r grab_snp_geno}
g <- do.call("cbind", svenson$geno[1:19])
fg <- do.call("cbind", svenson$founder_geno[1:19])
g <- g[,colSums(fg==0)==0]
fg <- fg[,colSums(fg==0)==0]
fgn <- colSums(fg==3)
```

We now calculate the genotype frequencies, by individual and by
marker. For the genotype frequencies by individual, we'll calculate
separate frequencies for the four groups of markers, split according
to their minor allele frequency in the founder strains.

```{r calc_snp_geno_freq, eval=FALSE}
gf_ind <- vector("list", 4)
for(i in 1:4) {
    gf_ind[[i]] <- t(apply(g[,fgn==i], 1, function(a) table(factor(a, 1:3))/sum(a != 0)))
}
```

```{r calc_snp_geno_freq_really, echo=FALSE}
file <- "diag_cache/snp_freq_ind.rds"
if(file.exists(file)) {
    gf_ind <- readRDS(file)
} else {
<<calc_snp_geno_freq>>
    saveRDS(gf_ind, file)
}
```


The following triangle plots show the genotype frequency
distributions for the mice, among the four groups of markers with
common minor allele frequency (MAF) in the founder strains. These
plots make use of the fact that for a point within an equilateral
triangle, the sum of the distances to the three sides is a constant.

```{r triplot_ind, fig.height=11}
par(mfrow=c(2,2), mar=c(0.6, 0.6, 2.6, 0.6))
for(i in 1:4) {
    triplot(c("AA", "AB", "BB"), main=paste0("MAF = ", i, "/8"))
    tripoints(gf_ind[[i]], pch=21, bg="lightblue")
    tripoints(c((1-i/8)^2, 2*i/8*(1-i/8), (i/8)^2), pch=21, bg="violetred")

    if(i>=3) { # label mouse with lowest het
      wh <- which(gf_ind[[i]][,2] == min(gf_ind[[i]][,2]))
      tritext(gf_ind[[i]][wh,,drop=FALSE] + c(0.02, -0.02, 0),
              names(wh), adj=c(0, 1))
    }

    # label other mice
    if(i==1) {
        lab <- rownames(gf_ind[[i]])[gf_ind[[i]][,2]>0.3]
    }
    else if(i==2) {
        lab <- rownames(gf_ind[[i]])[gf_ind[[i]][,2]>0.48]
    }
    else if(i==3) {
        lab <- rownames(gf_ind[[i]])[gf_ind[[i]][,2]>0.51]
    }
    else if(i==4) {
        lab <- rownames(gf_ind[[i]])[gf_ind[[i]][,2]>0.6]
    }

    for(ind in lab) {
        if(grepl("^F", ind) && i != 3) {
            tritext(gf_ind[[i]][ind,,drop=FALSE] + c(-0.01, 0, +0.01), ind, adj=c(1,0.5))
        } else {
            tritext(gf_ind[[i]][ind,,drop=FALSE] + c(0.01, 0, -0.01), ind, adj=c(0,0.5))
        }
    }
}
```

The majority of individuals are tightly clustered around the expected distribution
(in pink). There are a few outliers with elevated heterozygosity, but
these are all among the samples with highest rates of missing genotypes.
Mouse `r names(wh)` has slightly reduced heterozygosity
in markers with MAF=3/8 or 4/8, but not by a worrisome amount.


### Crossover counts

Counts of inferred crossovers can be a useful diagnostic for identifying
problem samples, which may show an excessive number of apparent crossovers.

We first reconstruct the haplotypes of the DO mice, at each position
calculating the probability of each of the 36 possible genotypes,
given the observed marker data.

```{r calc_genoprob, eval=FALSE}
pr <- calc_genoprob(svenson, error_prob=0.002, map_function="c-f", cores=0)
```

```{r calc_genoprob_really, echo=FALSE}
file <- "diag_cache/calc_genoprob.rds"
if(file.exists(file)) {
    pr <- readRDS(file)
} else {
<<calc_genoprob>>
    saveRDS(pr, file)
}
```

There are a couple of different ways to get estimates of the numbers
of crossovers. The simplest is to identify the most probable
genotype at each position, and then count exchanges in such data.
Here we use a cut-off of 0.5, leaving as missing any genotypes where
the largest probability is <0.5.

```{r maxmarg_countxo, eval=FALSE}
m <- maxmarg(pr, minprob=0.5, cores=0)
nxo <- count_xo(m, cores=0)
```

```{r maxmarg_countxo_really, echo=FALSE}
file <- "diag_cache/maxmarg.RData"
if(file.exists(file)) {
    load(file)
} else {
<<maxmarg_countxo>>
ph <- guess_phase(svenson, m)
    save(m, nxo, ph, file=file)
}
```


The function `count_xo()` returns counts of crossovers on each
chromosome (as columns) in each mouse. To get
estimates of the total numbers of crossovers, genome-wide, we take the
sums of each row:

```{r n_xo_maxmarg}
totxo <- rowSums(nxo)
```

Here's a plot of the number of crossovers vs the mouse index, colored by
generation. (We have omitted the `r sum(percent_missing >= 19.97)` samples
with > 20% missing genotypes, as these samples show
`r min(totxo[percent_missing >= 19.97])`&ndash;`r max(totxo[percent_missing >= 19.97])`
crossovers, well above the others.)

There is a lot of variation in the number of crossovers, and some overlap between
generation 8 (in red, with average `r round(totxo[svenson$covar$gen=="8" & percent_missing < 19.97])`
crossovers) and generation 11 (in blue, with average
`r round(totxo[svenson$covar$gen=="11" & percent_missing < 19.97])` crossovers).

```{r plot_nxo, fig.width=8.8, fig.height=4.8}
iplot(seq_along(totxo)[percent_missing < 19.97],
      totxo[percent_missing < 19.97],
      group=(svenson$covar$ngen=="11")[percent_missing < 19.97]+1,
      chartOpts=list(xlab="Mouse", ylab="Number of crossovers",
                     margin=list(left=80,top=40,right=40,bottom=40,inner=5),
                     axispos=list(xtitle=25,ytitle=50,xlabel=5,ylabel=5)))
```

We're looking for mice with an excessive number of
crossovers. Mice with very few crossovers would also deserve follow
up. And note that we need to take the DO generation of an individual
into account. But none of the mice seem unusual, in terms of numbers
of crossovers (except for the `r sum(percent_missing >= 19.97)` mice with >
20% missing genotypes, which we omitted from this figure.

Here are the crossover counts for those `r sum(percent_missing >= 19.97)`
mice:

```{r totxo_when_pmis_big}
tmp <- cbind(percent_missing=round(percent_missing), total_xo=totxo)[percent_missing >= 19.97,]
tmp[order(tmp[,1]),]
```



### Genotyping error LOD scores

[Lincoln and Lander
(1992)](https://doi.org/10.1016/S0888-7543(05)80158-2) proposed the
calculation of genotyping error LOD scores, to identify potential
genotyping errors. For each SNP in each mouse, we calculate a LOD
score statistic that measures the evidence that the particular
genotype is in error, by comparing the observed genotype to the
multipoint genotype probabilities, making use of the founder strains'
genotypes at the SNP. These are calculated using the function
`calc_errorlod()`.

```{r calc_errorlod, eval=FALSE}
e <- calc_errorlod(svenson, pr, cores=0)
```

```{r calc_errorlod_really, echo=FALSE}
file <- "diag_cache/errorlod.rds"
if(file.exists(file)) {
    e <- readRDS(file)
} else {
<<calc_errorlod>>
    saveRDS(e, file)
}
```

The output is a list of matrices (the components being chromosomes).
We can combine these into a single large matrix (mice &times; SNPs),
and then get estimated genotyping error rates for each mouse, taking
error LOD > 2 as a cutoff.

```{r plot_error_rate_ind, fig.width=8.8, fig.height=4.8}
e <- do.call("cbind", e)
errors_ind <- rowSums(e>2)/n_typed(svenson)*100
lab <- paste0(names(errors_ind), " (", myround(percent_missing,1), "%)")
iplot(seq_along(errors_ind), errors_ind, indID=lab,
      chartOpts=list(xlab="Mouse", ylab="Percent genotyping errors", ylim=c(0, 4.1),
                     axispos=list(xtitle=25, ytitle=50, xlabel=5, ylabel=5)))
```

There are `r sum(errors_ind > 1)` mice with estimated genotyping error
rates > 1%, and these are exactly the mice with > 20% missing
genotypes.
There are `r sum(errors_ind < 1 & errors_ind > 0.5)` mice with error
rates in the 0.5-1% range, and these are the mice with around 9%
missing genotypes.

The mean error rate of mice with <5% missing genotype data is
`r round(mean(errors_ind[percent_missing < 5])*100)` in 10,000.



### Apparent genotyping errors

A more direct approach to look for genotyping errors is to get
inferred SNP genotypes and compare them to the observed ones.
We use the result of `maxmarg()`, calculated above, and then
`predict_snpgeno()`. The output is again a list of matrices, one per
chromosome, but we'll combine them into a single matrix.

```{r predict_snpgeno}
snpg <- predict_snpgeno(svenson, m, cores=0)
snpg <- do.call("cbind", snpg)
```

We'll similarly grab the raw SNP genotypes.

```{r grab_raw_snps}
gobs <- do.call("cbind", svenson$geno)
gobs[gobs==0] <- NA
```

The apparent error rates derived by comparing the observed and
predicted SNP genotypes are basically the same as using a cutoff of 0
on the error LOD score.

```{r plot_error_rate_vs_elod0}
par(pty="s")
err_direct <- rowMeans(snpg != gobs, na.rm=TRUE)*100
errors_ind_0 <- rowSums(e > 0)/n_typed(svenson)*100
par(mar=c(4.1,4.1,0.6, 0.6))
grayplot(errors_ind_0, err_direct,
         xlab="Percent errors (error LOD > 0)",
         ylab="Percent errors (obs vs predicted)",
         xlim=c(0, 1.6), ylim=c(0, 1.6))
abline(0,1,lty=2, col="gray60")
```

Using a cutoff of 0 on the genotyping error LOD score, the average
error rate in mice with <5% missing genotypes is
`r round(mean(errors_ind_0[percent_missing < 5])*100)` per 10,000.

## Bad markers

Let's now turn to studying the markers, to find bad ones that we may
wish to omit.

We will first omit the mice with >20% missing genotype
data (which were also seen to have elevated numbers of apparent
crossovers and genotyping error rates).

```{r drop_high_missing}
n <- n_ind(svenson)
svenson <- svenson[percent_missing < 19.97,]
stopifnot( n_ind(svenson) == n-9 )

# update other stuff
e <- e[ind_ids(svenson),]
g <- g[ind_ids(svenson),]
```

### Missing data

It can also be useful to look at the proportion of missing genotypes
by marker. Markers with a lot of missing data were likely difficult to
call, and so the genotypes that _were_ called may contain a lot of
errors.

```{r prop_missing_marker}
pmis_mar <- n_missing(svenson, "marker", "proportion")*100
```

Here's a histogram of the proportion of missing genotypes by marker.
The tick marks below the histogram indicate individual values.

```{r hist_prop_missing_marker, dev="png", doi=600}
par(mar=c(5.1,0.6,0.6, 0.6))
hist(pmis_mar, breaks=seq(0, 100, length=201),
     main="", yaxt="n", ylab="", xlab="Percent missing genotypes")
rug(pmis_mar)
```

The vast majority of markers have very little missing data:
there are `r add_commas(sum(pmis_mar == 0))` markers
(`r round(mean(pmis_mar==0)*100)`%) that are missing _no_ data,
and `r add_commas(sum(pmis_mar < 2))` markers
(`r round(mean(pmis_mar<2)*100)`%) that are missing <2%.

But some markers are missing a lot of data, including
`r sum(pmis_mar > 50)` that are missing more than 50% of genotypes.
`r sum(pmis_mar > 30)` markers are missing more than 30% of
genotypes, while `r sum(pmis_mar > 15)` markers are missing more than
15% of genotypes. Recall that there are a total of `r add_commas(tot_mar(svenson))`,
markers, so `r sum(pmis_mar > 15)` markers is just
`r round(sum(pmis_mar > 15)/length(pmis_mar)*100, 1)`% of the total.

For the markers with lots of missing genotypes, it's not _necessarily_
the case that the remaining genotypes are full of errors, but in
studying the allele intensities at these SNPs, it does appear that for
the bulk of such markers, the genotypes are not being called
appropriately.


### Genotype frequencies

The genotype frequencies by marker can be informative, just as
genotype frequencies by individual were.

```{r calc_snp_freq_marker, eval=FALSE}
gf_mar <- t(apply(g, 2, function(a) table(factor(a, 1:3))/sum(a != 0)))
gn_mar <- t(apply(g, 2, function(a) table(factor(a, 1:3))))
```

```{r calc_snp_freq_marker_really, echo=FALSE}
file <- "diag_cache/snp_freq_marker.RData"
if(file.exists(file)) {
    load(file)
} else {
<<calc_snp_freq_marker>>
    save(gf_mar, gn_mar, file=file)
}
```


Here are scatter plots of the genotype frequencies by marker, split according to
the minor allele frequency in the 8 founders. A pink dot is placed at
the expected frequencies.

```{r triplot_bymar, fig.height=11}
par(mfrow=c(2,2), mar=c(0.6, 0.6, 2.6, 0.6))
for(i in 1:4) {
    triplot(c("AA", "AB", "BB"), main=paste0("MAF = ", i, "/8"))
    z <- gf_mar[fgn==i,]
    z <- z[rowSums(is.na(z)) < 3,]
    tripoints(z, pch=21, bg="gray80", cex=0.6)
    tripoints(c((1-i/8)^2, 2*i/8*(1-i/8), (i/8)^2), pch=21, bg="violetred")
}
```

The bulk of the markers seem well behaved, but there are a number of
markers with unusual genotype frequencies. There are
`r sum(gf_mar[,1]==0)` markers that show no homozygotes for the major allele.
(These sit on the left edge.) There are `r sum(gf_mar[,1]==1)` markers
that are monomorphic (100% AA genotypes; lower-right vertex). And there are
`r sum(gf_mar[,1]<1  & gf_mar[,2]==0)` markers with some of each
homozygote but no heterozygotes (bottom edge).

```{r calc_maf, echo=FALSE}
pB <- gf_mar[,3] + gf_mar[,2]*0.5
fg_mis <- colSums(fg==0)
```

Further, there are a bunch of markers where the minor allele appears
to be private to one founder strain (upper-left panel) but show a high
frequency of minor alleles in the DO offspring. For example, there are
`r sum(fgn==1 & fg_mis==0 & pB > 0.5)` markers where the minor
allele is private to one founder line but has >50% allele frequency in
the DO offspring, and `r sum(fgn==1 & fg_mis==0 & pB > 0.3)`
such private SNPs have >30% minor allele frequency in the DO
offspring. But it turns out that all but two of these are private
to WSB and reside on chromosome 2. This is due to meiotic drive of the
WSB allele at the R2D2 locus; see
[Didion et al. (2016)](https://doi.org/10.1093/molbev/msw036).

The other two are also interesting: JAX00463240 on chr 18
(private to CAST) and UNC9169851 on chr 5 (private to NZO) both have >30%
minor allele frequency in the DO offspring. Both seem to be in regions
where one of the founder alleles is unusually frequent.

### Genotyping errors

Let's look at the apparent genotyping errors by marker (again starting
with a cutoff of error LOD score > 2), plotted against the amount of
missing data.

```{r plot_errors_vs_missing}
errors_mar <- colSums(e>2)/n_typed(svenson, "marker")*100
grayplot(pmis_mar, errors_mar,
         xlab="Proportion missing", ylab="Proportion genotyping errors")
```

Most of the markers show no evidence for genotyping errors. Only
`r add_commas(sum(errors_mar > 0))`
(`r myround(sum(errors_mar>0)/tot_mar(svenson)*100,1)`% of markers)
show at least one apparent error, but
`r sum(errors_mar > 5)` show >5% errors. Markers with higher rates of
missing genotypes tend to show higher errors rates, though this is
difficult to detect in the above figure.



### Array intensities

Scatterplots of SNP allele intensities are quite revealing about the cause,
particularly when we look at the scatterplots colored by both the
observed SNP genotypes and by the predicted SNP genotypes, given the
multipoint SNP information. Here are a few examples.

Let us first use the inferred 36-state genotypes to get predicted SNP
genotypes, which we then combine into a single wide matrix.

```{r predict_snp_geno}
snpg <- predict_snpgeno(svenson, m, cores=0)
snpg <- do.call("cbind", snpg)
```

Here is a function to plot the SNP intensities, plus several helper functions.

```{r plot_intensities_function}
# get data for one marker
read_intensities <-
    function(file="../RawData/intensities.fst", marker=NULL, snps=NULL)
{
    if(is.null(snps)) {
        snps <- fst::read.fst(file, column="snp")[,1]
    }

    if(is.null(marker)) { # if marker is NULL, return the snp names
        return(snps)
    } else {
        marker <- as.character(marker)
    }

    n <- match(marker, snps)
    if(is.na(n)) stop("marker ", marker, " not found")

    reorg_intensities( fst::read.fst(file, from=n, to=n+1) )
}


# reorganize SNP intensity data (for one marker)
reorg_intensities <-
    function(intensities, marker=NULL)
{
    if(!is.null(marker)) {
        marker <- as.character(marker)
        wh <- which(intensities[,1] == marker)
        if(sum(wh==0)) stop("Marker ", marker, " not found")
        intensities <- intensities[wh,]
    }

    result <- data.frame(X=as.numeric(intensities[1,-(1:2)]),
                         Y=as.numeric(intensities[2,-(1:2)]))
    rownames(result) <- colnames(intensities)[-(1:2)]

    result
}

# combine intensities and genotypes
grab_gni <-
    function(marker=NULL, cross=svenson, intensities_file="../RawData/intensities.fst",
             intensities_data=NULL, drop_bad_samples=TRUE)
{
    if(is.null(intensities_data)) {
        # get intensities
        int <- read_intensities(intensities_file, marker=marker)
    } else {
        int <- reorg_intensities(intensities_data, marker)
    }
    marker <- as.character(marker)


    # get genotypes
    chr <- qtl2::find_markerpos(cross, marker)$chr
    g <- cross$geno[[chr]][,marker]
    gg <- setNames(rep(0, nrow(int)), rownames(int))
    gg[names(g)] <- g

    if(drop_bad_samples) {
        int <- int[names(g),,drop=FALSE]
        gg <- gg[names(g)]
    }

    cbind(int, g=gg)
}

# load intensities and plot them, colored by genotype calls
#
# drop_bad_samples: if TRUE, don't plot points for samples that are not in the cross object
#                   (e.g., were omitted previously as being bad DNAs)
#
plot_intensities <-
    function(marker, cross=svenson, intensities_file="diag_cache/intensities.fst",
             intensities_data=NULL, drop_bad_samples=TRUE, geno=NULL, ...)
{
    if(is.character(marker) || is.factor(marker)) {
        marker <- as.character(marker)
        gni <- grab_gni(marker, cross=cross, intensities_file=intensities_file,
                        intensities_data=intensities_data, drop_bad_samples=drop_bad_samples)
    } else {
        gni <- marker
    }

    if(!is.null(geno)) {
        if(is.logical(geno)) geno <- geno + 1 # FALSE/TRUE -> 1/2
        gni[names(geno),"g"] <- geno
    }

    internal_plot <-
        function(pch=21, bg=broman::brocolors("f2"),
                 xlab="allele 1", ylab="allele 2",
                 xlim=c(0, max(gni$X, na.rm=TRUE)),
                 ylim=c(0, max(gni$Y, na.rm=TRUE)),
                 ...)
        {
            grayplot(gni$X, gni$Y, pch=pch, bg=bg[match(gni$g, c(1:3,0))],
                     xlab=xlab, ylab=ylab,
                     xlim=xlim, ylim=ylim, ...)
        }

    internal_plot(...)

    invisible(gni)
}
```

The scatterplots below show examples of the allele intensities for
four SNP markers. Each point is a mouse sample, and the x and y axes
are for the array intensities of the two alleles at the SNP. In the
left panels, the points are colored by the observed SNP genotypes,
with yellow and blue corresponding to the two homozygotes and green to
the heterozygotes; gray points indicate missing genotypes. In the
right panels, the points are colored by the predicted SNP genotypes;
gray points indicate samples for which the SNP genotype could not be predicted.

```{r snp_intensities_examples, fig.height=15}
snpg[is.na(snpg)] <- 0

markers <- c("JAX00279019",
             "UNC12329705",
             "backupUNC140137407",
             "UNC20478577")

par(mfrow=c(4,2), mar=c(3.1,3.1,2.1,1.1))
for(i in seq_along(markers)) {
    mar <- markers[i]

    plot_intensities(mar, mgp.x=c(1.4,0.3,0), mgp.y=c(1.8,0.3,0))
    title(main=mar, line=1.05, cex.main=0.9)
    title(main="(observed genotypes)", cex.main=0.8, line=0.25)

    u <- par("usr")
    mtext(side=3, adj=0, LETTERS[i*2-1])

    plot_intensities(mar, geno=snpg[,mar],
                     mgp.x=c(1.4,0.3,0), mgp.y=c(1.8,0.3,0))
    title(main=mar, line=1.05, cex.main=0.9)
    title(main="(predicted genotypes)", cex.main=0.8, line=0.25)

    mtext(side=3, adj=0, LETTERS[i*2])
}
```

The top panels (A and B) are for a well-behaved SNP, `r markers[1]`.
There are three clear genotype clusters, and the observed and
predicted genotypes match.

In panels C and D (`r markers[2]`), the genotype
calling algorithm went awry. There are three distinct clusters of
samples, but the homozygotes for the major allele appear to have been
called heterozygotes.

In panels E and F (`r markers[3]`), there appears to be an extra
cluster of samples with distinct genotype, but the genotype calling
algorithm classified them as homozygotes and they instead (based on
surrounding markers) appear to be heterozygotes.

Panels G and H (`r markers[4]`) corresponds to a case where the
genotype clusters are overlapping and the genotyping calling algorithm
had difficulty distinguishing them.

## Effect of data cleaning

What's perhaps most important in data cleaning is to identify badly
behaved or mislabeled samples. We have identified a relatively small
number of markers that are showing appreciable genotyping errors, and
the worst of these should probably be omitted. Omitting poorly behaved
markers will have a small effect on later analysis results, as the HMM
to do genotype reconstructions smooths over such markers reasonably
well.

But it's worth omitting the bad markers and checking for differences
in the genotype probabilities.


Let's omit the `r sum(errors_mar > 5)` markers with error rates >5%.

```{r omit_error_prone_markers}
gmap <- svenson$gmap
pmap <- svenson$pmap
svenson <- drop_markers(svenson, names(errors_mar)[errors_mar > 5])
```

We'll now re-calculate the genotype probabilities.

```{r calc_genoprob_clean, eval=FALSE}
prcl <- calc_genoprob(svenson, gmap, error_prob=0.002, map_function="c-f", cores=0)
```

```{r calc_genoprob_clean_really, echo=FALSE}
file <- "diag_cache/calc_genoprob_clean.rds"
if(file.exists(file)) {
    prcl <- readRDS(file)
} else {
<<calc_genoprob_clean>>
    saveRDS(prcl, file)
}
```

As a measure of change, we'll use the sum of the absolute differences
between the probabilities, with and without the bad markers, at each
mouse and marker.

```{r pr_diff, eval=FALSE}
# make sure individual IDs are the same (omitting the 9 mice with >20% missing genotypes)
prcl <- prcl[ind_ids(svenson),]
pr <- pr[ind_ids(svenson),]

prdiff <- vector("list", length(pr))
for(i in seq_along(prdiff)) prdiff[[i]] <- apply(abs(pr[[i]] - prcl[[i]]), c(1,3), sum)
names(prdiff) <- names(pr)
```

```{r pr_diff_really, echo=FALSE}
file <- "diag_cache/prdiff.rds"
if(file.exists(file)) {
    prdiff <- readRDS(file)
} else {
<<pr_diff>>
    saveRDS(prdiff, file)
}

# make sure individual IDs are the same (omitting the 9 mice with >20% missing genotypes)
prcl <- prcl[ind_ids(svenson),]
pr <- pr[ind_ids(svenson),]
prdiff <- lapply(prdiff, function(a) a[ind_ids(svenson),])
```

These differences take values between 0 and 2, with a value of 2
meaning that completely different set of genotypes have positive
probability. Most chromosomes show some differences, but only at a
relatively sparse set of positions. Here are the number of markers
&times; individuals with differences > 1.5.

```{r sum_diff_gt_1}
sapply(prdiff, function(d) sum(d > 1.5))
```

And there are only a few individuals, on a given chromosome, showing
very many differences. Here are the number of individuals, by chromosome, with at least
give markers having absolute difference > 1.5.

```{r num_ind_with_10_diff_gt_1}
sapply(prdiff, function(d) sum(rowSums(d>1.5) > 5))
```

As an example of one of the changes, here is a bivariate heatmap of
the genotype probabilities for one chromosome in one individual. The
probabilities before cleaning are in blue, and the probabilities after
cleaning are in red, and then the two color scales are combined.
Probabilities that are high both before and after cleaning will be
dark puple. Probabilities that are high before cleaning but low after
cleaning will be blue, and probabilities that are low before cleaning
and high after cleaning will be red.

```{r plot_genoprobcomp_M305_c9}
plot_genoprobcomp(pr, prcl, pmap, ind="M305", chr="9", threshold=0.25)
```

So here, for individual M305 on chromosome 9, there is one region
at around 95 Mbp where, before data cleaning, the genotyped switched
from DE to BE and back, but after data cleaning it looks to be
constant DE.

The following figure shows the BE and DE probabilities for M305 on chr
9 in more detail.

```{r plot_specific_probs_m305_c9}
par(mfrow=c(2,1), mar=c(3.1, 3.1, 2.6, 0.6))
grayplot(pmap[[9]], pr[[9]]["M305","BE",], type="l",
         ylim=c(0,1), col="slateblue",
         xlab="Chr 9 position (Mbp)", ylab="Genotype probability",
         main="mouse M305\n(before cleaning)",
         mgp.x=c(1.4,0.3,0), mgp.y=c(1.9,0.3,0), lwd=2)
lines(pmap[[9]], pr[[9]]["M305","DE",], col="violetred", lwd=2)
legend("topright", lwd=2, col=c("slateblue", "violetred"),
       c("BE", "DE"), bg="gray92")

grayplot(pmap[[9]], prcl[[9]]["M305","BE",], type="l",
         ylim=c(0,1), col="slateblue",
         xlab="Chr 9 position (Mbp)", ylab="Genotype probability",
         main="mouse M305\n(after cleaning)",
         mgp.x=c(1.4,0.3,0), mgp.y=c(1.9,0.3,0), lwd=2)
lines(pmap[[9]], prcl[[9]]["M305","DE",], col="violetred", lwd=2)
legend("topright", lwd=2, col=c("slateblue", "violetred"),
       c("BE", "DE"), bg="gray92")
```


Omitting poorly behaved markers can affect the genotype probabilities,
but these effects appear to be rather isolated.


## Session info

Here are the versions of R and packages that I'm using in this
analysis.

```{r session_info}
devtools::session_info()
```
