---
title: Benchmarks of HMM routines
output: rmarkdown::html_vignette
---

```{r change_width, include=FALSE}
options(width=100)
```

[R/qtl2geno](https://github.com/kbroman/qtl2geno) includes a reimplementation
of the basic hidden Markov model code for the calculation of
conditional genotype probabilities (given the available multipoint
marker genotype data) and the estimation of genetic maps.

The new HMM code can potentially handle more general cases and uses
C++ classes to simplify the extension to new cross types. I've also
split out the forward and backward equations as separate functions, to
reduce code duplication. And the basic calculations are separated by
individual, which may allow for low-level parallel processing (i.e.,
[concurrency](https://en.wikipedia.org/wiki/Concurrency_(computer_science))).
(I'm considered using
[RcppParallel](http://dirk.eddelbuettel.com/blog/2014/07/16/#introducing_rcppparallel)
for this.)

The new code for calculating conditional genotype probabilities is
about the same as that in [R/qtl](http://www.rqtl.org).  The new code
for estimating the genetic map, on the other hand, is considerably
slower than that in [R/qtl](http://www.rqtl.org). The present document
provides detailed benchmarks on this point.

## Genotype probabilities

First let's look at the calculation of conditional genotype
probabilities (given multiple genetic marker genotypes).  We'll
consider simulated data, first of a backcross with 500 individuals
(250 males and 250 females) and with markers at a 1 cM spacing on 20
chromosomes (including the X chromosome), and 5% missing data. We use
`qtl::sim.map()` and `qtl:sim.cross()` to simulate the data.

```{r sim_backcross}
set.seed(8440561)
library(qtl)
data(map10)
chrL <- round(summary(map10)$length[1:20])
map <- sim.map(len=chrL, n.mar=chrL+1, include.x=TRUE, eq.spacing=TRUE)
n <- 500
bc <- sim.cross(map, n.ind=n, type="bc", missing.prob=0.05)
bc$pheno$sex <- rep(0:1, each=n/2)
```

There's a total of `r totmar(bc)` markers.

We use `qtl2geno::convert2cross2()` to convert the data to the format for
[R/qtl2](http://kbroman.org/qtl2).

```{r convert2cross2}
library(qtl2geno)
bc2 <- convert2cross2(bc)
```

We use the
[microbenchmark package](https://cran.r-project.org/package=microbenchmark)
and run each of `qtl::calc.genoprob()` and `qtl2geno::calc_genoprob2()`
25 times, with calculations at a 0.5 cM spacing. Even though the data
were simulated without genotyping errors, I'm going to do the
calculations assuming a 0.2% genotype error rate.

```{r benchmark_calcgenoprob_bc}
library(microbenchmark)
probbc <- microbenchmark(qtl =  qtl::calc.genoprob(bc, step=0.5, error.prob=0.002),
                         qtl2 = qtl2geno::calc_genoprob(bc2, step=0.5, error_prob=0.002),
                         times=25)
print(probbc, unit="s")
```

```{r median_ratio, include=FALSE}
library(broman)
library(magrittr)
# get ratio of median times
median_ratio <-
function(mbmout,digits=1)
  summary(mbmout)$median %>% log %>% diff %>% exp %>% myround(., digits)
```

Let's do the same for an intercross, also including the X chromosome,
and with crosses in both directions and with both sexes.

```{r sim_intercross}
f2 <- sim.cross(map, n.ind=n, type="f2", missing.prob=0.05)
f2$pheno$sex <- rep(0:1, each=n/2)
f2$pheno$pgm <- rep(0:1, n/2)
f2_2 <- convert2cross2(f2)
```

Now for the benchmarks; we'll do the calculations just 10
times.

```{r benchmark_calcgenoprob_f2}
probf2 <- microbenchmark(qtl =  qtl::calc.genoprob(f2, step=0.5, error.prob=0.002),
                         qtl2 = qtl2geno::calc_genoprob(f2_2, step=0.5, error_prob=0.002),
                         times=10)
print(probf2, unit="s")
```


## Genetic map estimation

Now let's consider the estimation of inter-marker distances in the
genetic map, with the functions `qtl::est.map` and `qtl2geno::est_map`.

First the backcross; to speed things up, we'll look at just
chromosomes 1 and X.

```{r bc_subset_chr}
bc <- bc[c(1, "X"),]
bc2 <- convert2cross2(bc)
```

Now for the benchmarks. We'll just do 5 replicates.

```{r benchmark_estmap_bc}
mapbc <- microbenchmark(qtl =  qtl::est.map(bc, error.prob=0.002, tol=1e-6, maxit=10000),
                        qtl2 = qtl2geno::est_map(bc2, error_prob=0.002, tol=1e-6, maxit=10000),
                        times=5)
print(mapbc, unit="s")
```

Now to the intercross. Again, we'll look at just
chromosomes 1 and X, and we'll perform just 5 replicates.

```{r benchmark_estmap_f2}
f2 <- f2[c(1, "X"),]
f2_2 <- convert2cross2(f2)
mapf2 <- microbenchmark(qtl =  qtl::est.map(f2, error.prob=0.002, tol=1e-6, maxit=10000),
                        qtl2 = qtl2geno::est_map(f2_2, error_prob=0.002, tol=1e-6, maxit=10000),
                        times=5)
print(mapf2, unit="s")
```

For the backcross, the new code takes `r median_ratio(mapbc)` times
longer; for the intercross, the new code takes
`r median_ratio(mapf2)` times longer.


## Why so slow?

I made a number of changes to the HMM code between
[R/qtl](http://www.rqtl.org) and [R/qtl2](http://kbroman.org/qtl2).

- Split out the forward/backward equations as separate functions (to
  avoid code duplication and to make the code potentially clearer).
- Calculations are by individual, with an eye towards low-level
  parallel processing
- Things are a bit more general. For example, the transition
  probabilities in the Markov chain can be completely different for
  each individual, depending on the `cross_info` data. Also, the
  estimation of recombination fractions is potentially cross-specific;
  this allows us to handle RIL with the same basic code as a backcross.
- Using R-style matrix objects (via [Rcpp](http://rcpp.org)) rather
  than low-level C-type arrays.

We could just rely on parallel processing to speed things up. (I've
implemented parallel calculations at the level of chromosomes, using
the
[parallel](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf)
package, but we could make more efficient use of
available processors by splitting things out at a lower level.) Other
alternatives are to have separate implementations for different groups
of crosses, or to identify the set of observed `cross_info` values and
pre-calculate emission and transition probabilities for each.

It's also possible that some basic aspects of my code could be
improved, even keeping the overall design fixed. I'm new to C++ and
not entirely sure that I'm doing things in the best way.


## Session information

The following shows the R and package versions I was using.

```{r sessionInfo, results="markup"}
devtools::session_info()
```
