# R/qtl2: Benchmarks of HMM routines

[R/qtl2](https://github.com/kbroman/qtl2) includes a reimplementation
of the basic hidden Markov model code for the calculation of
conditional genotype probabilities (given the available multipoint
marker genotype data) and the estimation of genetic maps.

The new HMM code can potentially handle more general cases and uses
C++ classes to simplify the extension to new cross types. I've also
split out the forward and backward equations as separate functions, to
reduce code duplication. And the basic calculations are separated by
individual, which may allow for low-level parallel processing (i.e.,
[concurrency](http://en.wikipedia.org/wiki/Concurrency_(computer_science))).
(I'm considered using
[RcppParallel](http://dirk.eddelbuettel.com/blog/2014/07/16/#introducing_rcppparallel)
for this.)

_However_, the current code is showing pretty crappy performance
relative to [R/qtl](http://www.rqtl.org), and may need to be
rethought. For example, `qtl2::calc_genoprob()` is about twice as slow
as `qtl::calc.genoprob()`, and `qtl2::est_map()` is about six times as
slow as `qtl::est.map()`. The present document provides detailed
benchmarks on this point.

## Genotype probabilities

First let's look at the calculation of conditional genotype
probabilities (given multiple genetic marker genotypes).  We'll
consider simulated data, first of a backcross with 500 individuals
(250 males and 250 females) and with markers at a 1 cM spacing on 20
chromosomes (including the X chromosome), and 5% missing data. We use
`qtl::sim.map()` and `qtl:sim.cross()` to simulate the data.

```{r sim_backcross}
set.seed(8440561)
library(qtl)
data(map10)
chrL <- round(summary(map10)$length[1:20])
map <- sim.map(len=chrL, n.mar=chrL+1,
               include.x=TRUE, eq.spacing=TRUE)
n <- 500
bc <- sim.cross(map, n.ind=n, type="bc", missing.prob=0.05)
bc$pheno$sex <- rep(0:1, each=n/2)
```

There's a total of `r totmar(bc)` markers.

We use `qtl2::convert2cross2()` to convert the data to the format for 
[R/qtl2](http://kbroman.org/qtl2).

```{r convert2cross2}
library(qtl2)
bc2 <- convert2cross2(bc)
```

We use the
[microbenchmark package](http://cran.r-project.org/web/packages/microbenchmark/)
and run each of `qtl::calc.genoprob()` and `qtl2::calc_genoprob2()`
25 times, with calculations at a 0.5 cM spacing. Even though the data
were simulated without genotyping errors, I'm going to do the
calculations assuming a 0.2% genotype error rate.

```{r benchmark_calcgenoprob_bc}
library(microbenchmark)
microbenchmark(qtl =  qtl::calc.genoprob(bc, step=0.5, error.prob=0.002),
               qtl2 = qtl2::calc_genoprob(bc2, step=0.5, error_prob=0.002),
               times=25)
```

Let's do the same for an intercross, also including the X chromosome,
and with crosses in both directions and with both sexes.

```{r sim_intercross}
f2 <- sim.cross(map, n.ind=n, type="f2", missing.prob=0.05)
f2$pheno$sex <- rep(0:1, each=n/2)
f2$pheno$pgm <- rep(0:1, n/2)
f2_2 <- convert2cross2(f2)
```

Now for the benchmarks; we'll do the calculations just 10
times.

```{r benchmark_calcgenoprob_f2}
microbenchmark(qtl =  qtl::calc.genoprob(f2, step=0.5, error.prob=0.002),
               qtl2 = qtl2::calc_genoprob(f2_2, step=0.5, error_prob=0.002),
               times=10)
```


## Genetic map estimation

Now let's consider the estimation of inter-marker distances in the
genetic map, with the functions `qtl::est.map` and `qtl2::est_map`.

First the backcross; to speed things up, we'll look at just
chromosomes 1 and X.

```{r bc_subset_chr}
bc <- bc[c(1, "X"),]
bc2 <- convert2cross2(bc)
```

Now for the benchmarks. We'll just do 5 replicates.

```{r benchmark_estmap_bc}
microbenchmark(qtl =  qtl::est.map(bc, error.prob=0.002,
                                   tol=1e-6, maxit=10000),
               qtl2 = qtl2::est_map(bc2, error_prob=0.002,
                                    tol=1e-6, maxit=10000),
               times=5)
```

Now to the intercross. Again, we'll look at just
chromosomes 1 and X, and we'll perform just 5 replicates.

```{r benchmark_estmap_f2}
f2 <- f2[c(1, "X"),]
f2_2 <- convert2cross2(f2)
microbenchmark(qtl =  qtl::est.map(f2, error.prob=0.002,
                                   tol=1e-6, maxit=10000),
               qtl2 = qtl2::est_map(f2_2, error_prob=0.002,
                                    tol=1e-6, maxit=10000),
               times=5)
```


## Session information

The following shows the R and package versions I was using.

```{r sessionInfo, results="markup"}
devtools::session_info()
```

<!-- the following to make it look nicer -->
<link href="http://kbroman.org/qtlcharts/assets/vignettes/vignette.css" rel="stylesheet"></link>
