---
title: R/qtl2 input file format
output: rmarkdown::html_vignette
---

<meta name="description" content="Input file format for R/qtl2">
<meta name="author" content="Karl Broman">

[R/qtl2](http://kbroman.org/qtl2) (aka qtl2) is a reimplementation of the QTL analysis software
[R/qtl](https://rqtl.org), to better handle high-dimensional data
and complex cross designs.

The input data file formats for [R/qtl](https://rqtl.org) cannot
handle complex crosses, and so for R/qtl2, we need to define a new
input file format. This document describes the details.

For simple cross types, we can continue to use the file formats for
[R/qtl](https://rqtl.org), use `qtl::read.cross()` to read in the
data, and then use a conversion function (`qtl2geno::convert2cross2()`) to convert
the data into the new format.

For more complex crosses, we need to define a new format. I was
persuaded by [Aaron Wolen](http://aaronwolen.com/)'s idea of a
[&ldquo;tidy&rdquo; format](https://github.com/kbroman/qtl/pull/20)
for [R/qtl](https://rqtl.org), with three separate CSV files, one
for phenotypes, one for genotypes, and one for the genetic map.

Another important idea is from [Pjotr Prins](http://thebird.nl/)'s
[qtab format](https://github.com/qtlHD/qtlHD/blob/master/doc/input/qtab.md):
the inclusion of metadata, such as genotype
encodings, with the primary data. This will simplify the handling of
multiple files and will help to avoid mistakes.

And so the basic idea for the new format is to have a separate file
for each part of the primary data (genotypes, founder genotypes,
genetic map, physical map, phenotypes, covariates, and phenotype
covariates), and then a control file (in [YAML](http://www.yaml.org)
or [JSON](http://www.json.org) format) which specifies the names of all
of those files, the genotype encodings and missing value codes, and
things like the name of the sex column within the covariate data (and
the encodings for the sexes) and which chromosome is the X chromosome.

## Reading the data files

Before discussing the boring file specifications, let's consider briefly
how the data are read into R.

A key advantage of the control file scheme is that it greatly
simplifies the function for reading in the data. That function,
`read_cross2()`, has a _single_ argument: the name (with path) of the
control file. So you can read in data like this:

```{r read_cross2, eval=FALSE}
library(qtl2geno)
grav2 <- read_cross2("~/my_data/grav2.yaml")
```

The large number of files is a bit cumbersome, so we've made it
possible to use a
[zip file](https://en.wikipedia.org/wiki/Zip_(file_format)) containing
all of the data files, and to read that zip file directly. There's even a
function for creating the zip file:

```{r create_zip, eval=FALSE}
zip_datafiles("~/my_data/grav2.yaml")
```

The `zip_datafiles()` function will read the control file to identify
all of the relevant data files and then zip them up into a file with
the same name and location, but with the extension `.zip` rather than
`.yaml` or `.json`.

To read the data back in, we use the same `read_cross2()` function,
providing the name (and path) of the zip file rather than the control
file.

```{r read_zip, eval=FALSE}
grav2 <- read_cross2("~/my_data/grav2.zip")
```

This can even be done with remote files.

```{r read_remote_zip, eval=FALSE}
grav2 <- read_cross2("http://kbroman.org/qtl2/assets/sampledata/grav2/grav2.zip")
```

Of course, the other advantage of the zip file is that it is
_compressed_ and so smaller than the combined set of CSV files.

## Format of the data files

The bulk of the data is in a set of comma-delimited (CSV) files. In
addition, a control file (in [YAML](http://www.yaml.org)
or [JSON](http://www.json.org) format),
contained in the same directory as the CSV files,
specifies the file names and other control parameters (such as
genotype and sex encodings). Sample data files are available at
[the R/qtl2 website](http://kbroman.org/qtl2/pages/sampledata.html).
We'll discuss the CSV files first.


### CSV files

The comma-delimited (CSV)
files are each in the form of a simple matrix, with the first
column being a set of IDs and the first row being a set of variable names.

Missing value codes will be specified in the control file (as
`na.strings`, with default value `"NA"`) and will apply across all files, so a missing value
code for one file cannot be an allowed value in another file.

The CSV files can include a header with a set of comment lines
initiated by a value specified in the control file as `comment.char`
(with default value `"#"`). The first such line could be a description
of the contents of the file. These comment lines can include the
expected number of rows and columns, like this:

    # This file contains blah, blah, blah...
    # nrow 25012
    # ncol 91

The number of rows (`nrow`) includes *only* the data rows (*not* the
comment rows, *nor* the row with variable names). On the other hand, the
number of columns (`ncol`) *does include* the column with individual
IDs.

All of these CSV files may be transposed relative to the form described
below. You just need to include, in the control file, a line like

    geno_transposed: true


#### Genotype and founder genotype data

The genotype data file is a matrix of individuals &times; markers. The first
column is the individual IDs; the first row is the marker names. The founder
genotypes (if needed) are in the same form, with founder lines as rows
and markers as columns, and with founder IDs in the first column.

#### Phenotype and covariate data

We split the numeric phenotypes from the mixed-mode covariates, as two
separate CSV files. Each file forms a matrix of individuals &times;
phenotypes (or covariates), with the first column being individual IDs
and the first row being phenotype or covariate names. Sex and line IDs
(if needed) can be columns in the covariate data.

#### Phenotype covariates

A separate CSV file contains phenotype covariate data, as phenotypes
&times; phenotype covariates. The first column
contains phenotype names, and the first row contains the names of the
phenotype covariates.

**Note**: The genotype, founder genotype, phenotype, covariate, and
  phenotype covariate data can be split across multiple files. For
  example, the genotype data could be split by chromosome. The
  individual IDs must appear in each file; these are used to combine
  the files.

#### Genetic and physical maps

Genetic and physical maps of the genotyped markers will be as separate
CSV files, each with three columns: marker, chromosome, and
position. The first row should be `marker,chr,pos` but will be
ignored. In the genetic map file, positions should be in centiMorgans
(cM). In the physical map file, positions should be in megabasepairs
(Mbp).


#### Cross information

The `"cross_info"` data specifies details of the cross that generated
each individual and is a numeric matrix with individuals as rows
(the same number of rows as in the genotype data) and with columns
depending on the cross type.

For simple cross types (e.g., `"f2"`, an intercross between two inbred
lines), this cross information may be included as a column in the
covariate data. More generally, the cross information will be a
separate CSV file. For example, for a set of Collaborative Cross (CC)
lines, we will want a matrix with eight columns, which indicate the
order of the founders in the crosses that generated each CC line.

So, in general, the cross information will be in a CSV file with individuals as
rows and a set of columns that define the cross information for that
cross type. The first column contains individual IDs and the first row
contains column names. Details on the column information are provided
in the cross-type-specific information, below.


### Control file

The new input file format includes a text-based control file (in
[YAML](http://www.yaml.org) or [JSON](http://www.json.org) format) to specify the names of all of the other
files as well as various control parameters such as genotype and sex
encodings and codes for missing values. We use
[YAML](http://www.yaml.org) because it is flexible, readable,
and easy to import into R. We also allow [JSON](http://www.json.org);
though it is often less human-readable, it can be less prone to errors.

The format of the control file is a bit technical. We
describe the details here and also provide a function
`write_control_file()` that takes the detailed specifications as input
and contructs the control file in the correct format.

We'll start with an example: the control file for the
[sample intercross data](http://kbroman.org/qtl2/pages/sampledata).

```
# Data from Grant et al. (2006) Hepatology 44:174-185
# Abstract of paper at PubMed: https://www.ncbi.nlm.nih.gov/pubmed/16799992
# Available as part of R/qtl book package, https://github.com/kbroman/qtlbook
crosstype: f2
geno: iron_geno.csv
pheno: iron_pheno.csv
phenocovar: iron_phenocovar.csv
covar: iron_covar.csv
gmap: iron_gmap.csv
alleles:
- S
- B
genotypes:
  SS: 1
  SB: 2
  BB: 3
sex:
  covar: sex
  f: female
  m: male
cross_info:
  covar: cross_direction
  (SxB)x(SxB): 0
  (BxS)x(BxS): 1
x_chr: X
na.strings:
- '-'
- NA
```

Any line that begins with a &ldquo;`#`&rdquo; is treated as a comment and
ignored. It's good to include some comments at the top of the file, describing the
dataset.

The order of things within the file is not important, but the names
of things are critical.

Much of the information is represented as key-value pairs, as
&ldquo;`key: value`.&rdquo;  For example, the cross type is indicated with a line like

```
crosstype: f2
```

The &ldquo;`key`&rdquo; is &ldquo;`crosstype`&rdquo; and the
&ldquo;`value`&rdquo; is &ldquo;`f2`.&rdquo; This indicates that the
data are for an F<sub>2</sub> intercross between two inbred lines.


#### File names

The names of the basic CSV files are indicated with lines like

```
geno: iron_geno.csv
```

This indicates that the genotype data are in the file `iron_geno.csv`.
The files are expected to be in the same directory as the control
file. They could be placed in separate directories, with the file names
being paths relative to the location of the control file, but this
is not recommended (or well tested).

The &ldquo;keys&rdquo; for the different files are the following:

```
geno:         genotype_filename
founder_geno: founder_genotype_filename
pheno:        phenotype_filename
covar:        covariate_filename
phenocovar:   phenotype_covariate_filename
gmap:         genetic_map_filename
pmap:         physical_map_filename
```

Most of these files are optional; if a particular file is not used,
the corresponding key can be omitted from the control file.

If the data for a section is split into multiple files (for example,
if the genotypes are split into chromosome-specific files), then a
vector of file names should be provided. For example:

```
geno:
  - geno1.csv
  - geno2.csv
  - genoX.csv
founder_geno:
  - founder_geno1.csv
  - founder_geno2.csv
  - founder_genoX.csv
```

#### X chromosome

If one of the chromosomes is to be treated as the X chromosome, there
should be a line like

```
x_chr: X
```

This specifies the chromosome ID for the X chromosome (`X` in this case).

#### Allele labels

To add labels in summary tables and plots, provide a vector of
single-character allele labels, with one for each founder line.
For example,

```
alleles:
- S
- B
```

This list of items, each beginning with a hyphen and a space, is the
YAML format for a vector. It is equivalent to the R code `c("S", "B")`.

You could also write this line as

```
alleles: [S, B]
```

which is an alternative format for vectors in YAML.


#### Genotype codes

The control file should contain a record with &ldquo;`genotypes:`&rdquo; that specifies
the genotype encodings. Here's an example:

```
genotypes:
  SS: 1
  SB: 2
  BB: 3
```

For each possible genotype code, indent and provide a &ldquo;`key: value`&rdquo;
pair, with the key being the code used in the genotype and founder
genotype files, and the value being an integer to which the genotype
should be converted.

The above example would be suitable for a backcross or intercross. For
a backcross, the second homozygote (`BB` in this case) is only needed
in the case that there are X chromosome genotypes for males.

For RIL, we would use something like

```
genotypes:
  BB: 1
  DD: 2
```

For crosses with multiple parents, the genotype file should contain
genotype calls for a set of SNPs, and there should be a corresponding
founder genotype file with genotypes of the founders at those SNPs.
A common set of genotype codes needs to be used for all SNPs.
In particular, the genotypes **cannot** be encoded as `AA`, `CC`,
`GG`, `TT`, `AC`, `AG`, because then, e.g., `CC` would need to be
treated as `1` for some SNPs and `3` for others. Instead, code the
genotypes with something like `AA`, `AB`, `BB`, and then include the
following in the control file:

```
genotypes:
  AA: 1
  AB: 2
  BB: 3
```

#### Sex

Sex can be provided as a column in the covariate file or as a separate
file.

If it is a column in the covariate file, the control file should have
a section that looks like this:

```
sex:
  covar: sex
  f: female
  m: male
```

Here, &ldquo;`covar: sex`&rdquo; indicates that the column name used in the
covariate file is &ldquo;`sex`.&rdquo; If the column name were &ldquo;`Sex`,&rdquo; you would
write &ldquo;`covar: Sex`.&rdquo;

The other two &ldquo;`key: value`&rdquo; pairs are the encodings used for
sex, with the &ldquo;keys&rdquo; being the codes used in the covariate file and the
&ldquo;values&rdquo; being `female` and `male`. So this indicates that sex was
encoded as `f` for females and `m` for males. If, instead, the sex
covariate had `0` for females and `1` for males, you would use:

```
sex:
  covar: sex
  0: female
  1: male
```

Sex information can also be provided as a separate file. In this case,
the file should have two columns: individual ID, and sex. Further, the
part of the control file dealing with sex should look like this:

```
sex:
  file: sex_filename
  f: female
  m: male
```

So instead of a line with &ldquo;`covar:`,&rdquo; use &ldquo;`file:`&rdquo; followed by the name
of the file (e.g., &ldquo;`file: iron_sex.csv`&rdquo;). You must still provide the sex
encodings, as before.


#### Cross information

For simple crosses (e.g., an intercross), cross information can be a
single column within the covariate file. In this case, include
something like the following in the control file:

```
cross_info:
  covar: cross_direction
  (SxB)x(SxB): 0
  (BxS)x(BxS): 1
```

This is much like the information for sex. The &ldquo;`covar:`&rdquo; line
indicates the name of the column in the covariate data that
corresponds to the cross information. The other two lines indicate
the encodings of the cross information as &ldquo;`key: value`&rdquo;
pairs, where &ldquo;`key`&rdquo; is
the code used in the cross information column and &ldquo;`value`&rdquo; is the
integer to which it should be converted.

More generally, the cross information would be contained in a separate
comma-delimited file. For simple crosses, in which the cross
information is a single column, we allow it to be encoded differently
from what is needed, and the control file information should look like
this:

```
cross_info:
  file: crossinfo_filename
  (SxB)x(SxB): 0
  (BxS)x(BxS): 1
```

For more complex crosses (e.g., the Collaborative Cross), the cross
information spans multiple columns and we require that the user have
set this up in advance (i.e., no translation of encodings will be
performed). In this case the relevant section of the control file
looks like this:

```
cross_info:
  file: crossinfo_filename
```

Or, more simply, you could write:

```
cross_info: crossinfo_filename
```

#### Missing value codes

To indicate the set of codes that are to be treated as missing values
in the genotype, founder genotype, phenotype, covariate, and phenotype
covariate files, define `na.strings` within the control file:

```
na.strings:
- NA
- '-'
```

A hyphen needs to be surrounded in single- or double-quotes. Many
other character strings (such as `NA`) do not. This is a similar
contruction as for the allele codes above; the list with hyphens
followed a space is the YAML format for a vector. You could also
write:

```
na.strings: [NA, '-']
```

which is another way to define a vector with YAML.

#### Field separator

If the data files use a separator other than a comma (e.g., a
semi-colon, or the vertical bar (`|`) which I like because it is seldom
present in data), indicate the separator within the control file, as
follows:

```
sep: '|'
```

A vertical bar needs to be surrounded by single- or double-quotes. A
semicolon doesn't, but it doesn't hurt if you do.





## Detailed specifications for each cross type

### Backcross

The cross type for a backcross is `"bc"`. In the case that the X
chromosome is included, a sex covariate needs to be included. The
genotype codes are `1` for homozygotes and `2` for heterozygotes, and
so the yaml control file will contain lines like the following:

```
genotypes:
  AA: 1
  AB: 2
```

The values on the left-hand side are the genotype codes in the
genotype data file; the numbers on the right are the numeric codes
they'll be converted into.

The equivalent genotype code specification for a json control file is

```
"genotypes": {"AA": 1, "AB": 2}
```


### F~2~ intercross

The cross type for an intercross between two inbred strains is `"f2"`.

The genotypes will be encoded as `1`, `2`, and `3` for say AA, AB, and
BB. The yaml control file will contain lines like the following:

```
genotypes:
  AA: 1
  AB: 2
  BB: 3
```

The values on the left-hand side are the genotype codes in the
genotype data file; the numbers on the right are the numeric codes
they'll be converted into. The equivalent JSON specification is:

```
"genotypes": {"AA": 1, "AB": 2, "BB": 3}
```

If the X chromosome is included in the data, we will need a sex
covariate as well as cross information. The cross information for an
intercross indicates the direction of the cross, with `0`
corresponding to the forward direction, for example
(A&times;B)&times(A&times;B) with females listed first, and `1`
corresponding to the reverse direction, for example
(B&times;A)&times;(B&times;A). What matters is the direction for the
cross that led to the males, since the F~2~ females get an intact X
chromosome from their father, which is A for the forward direction and
B for the reverse. Different individuals can come from different
directions, and you indicate the encodings with lines like the
following:

```
cross_info:
  covar: cross_direction
  (AxB)x(AxB): 0
  (BxA)x(BxA): 1
```

Here, we're saying that the covariate file has a column named
"`cross_direction`" that contains the cross information, and that the
forward direction is encoded as `(AxB)x(AxB)` and the reverse
direction as `(BxA)x(BxA)`.

You can have multiple codes for the cross direction information. For
example, the `cross_direction` covariate might use four possible
codes, including not just `(AxB)x(AxB)` and `(BxA)x(BxA)`, but also
`(BxA)x(AxB)`, and `(AxB)x(BxA)`.

In this case, the YAML control file should contain the following:

```
cross_info:
  covar: cross_direction
  (AxB)x(AxB): 0
  (BxA)x(AxB): 0
  (BxA)x(BxA): 1
  (AxB)x(BxA): 1
```

The equivalent JSON specification is:

```
"cross_info":{"covar":"cross_direction", "(AxB)x(AxB)":0, "(BxA)x(AxB)":0, "(BxA)x(BxA)":1, "(AxB)x(BxA)":1}
```

Here are some example data files for F~2~ intercrosses, in R/qtl2
format:

- [`iron`](http://kbroman.org/qtl2/assets/sampledata/iron/)
  from [Grant et al. (2006) Hepatology 44:174-185](https://www.ncbi.nlm.nih.gov/pubmed/16799992).
  A [zip file for these data](http://kbroman.org/qtl2/assets/sampledata/iron/iron.zip) is distributed
  with the [R/qtl2geno](https://github.com/rqtl/qtl2geno) package.

- [`B6BTBR`](https://github.com/rqtl/qtl2data/tree/master/B6BTBR) from
  [Tian et al. (2015) Genetics 201:1253-1262](https://doi.org/10.1534/genetics.115.179432)
  and [Broman et al. (2015) G3 5:2177-2186](https://doi.org/10.1534/g3.115.019778).

- [`Gough`](https://github.com/rqtl/qtl2data/tree/master/Gough) from
  [Gray et al. (2015) Genetics 201:213-228](https://doi.org/10.1534/genetics.115.177790).


### Recombinant inbred lines (RIL)

The cross types for a recombinant inbred lines derived from two inbred
strains are `"riself"` and
`"risib"` for RIL by selfing and sib-mating, respectively.

### Doubled haploids (DH)

The cross type for doubled haploids is `"dh"`.

### Haploids

The cross type for haploids is `"haploid"`.

### Advanced intercross lines (AIL)

The cross type for advanced intercross lines derived from two inbred
strains is `"ail"`.

### Diversity Outbreds (DO)

The cross type for the Diversity Outbred mouse population is `"do"`.

### Heterogeneous stock (HS)

The cross type for the heterogenous stock derived from 8 inbred
strains is `"hs"`.


### Multi-parent recombinant inbred lines

R/qtl2 can handle multi-parent recombinant inbred lines by selfing in
the case of 4, 8, or 16 founder lines (`"riself4"`, `"riself8"`, and
`"riself16"`), and by sib-mating in the case of 4 or 8 founder lines
(`"risib4"` and `"risib8"`). The Collaborative Cross (CC) is an
example of 8-way RIL by sib-mating (`"risib8"`). Many MAGIC
populations are also examples of multi-way RIL by sibling mating.

Two related cross types are `"magic19"` and `"dh6"`. The `"magic19"`
cross type is specific for the 19-way Arabidopsis MAGIC lines of
[Kover et al (2009) PLOS Genet 5:e1000551 doi:10.1371/journal.pgen.1000551](https://doi.org/10.1371/journal.pgen.1000551).
This was derived from a full diallel (in both directions) among 19
strains followed by 3 generations of random mating (to F4) and then
selfing for 6 generations. The `"dh6"` cross type is for a set of
maize MAGIC populations developed at the Wisconsin Crop Innovation
Center: A diallel of 6 founders followed by random mating for some
number of generations and then derivation of doubled haploids.
