# R/qtl2: Benchmarks of linear regression routines

[R/qtl2](https://github.com/kbroman/qtl2) includes code with several
options for linear regression. We're mostly focusing on using the
[Eigen library](http://eigen.tuxfamily.org) via
[RcppEigen](https://github.com/RcppCore/RcppEigen), particularly
linear regression via QR decomposition with column pivoting. We want
speed, but we often have to deal with rank-deficient matrices.

In addition to routines that return coefficient estimates and
estimated standard errors (SEs), we also have routines that return
just the residual sum of squares (RSS), as for the calculation of LOD
scores, that's all that matters. We might perform a lot of
regressions, getting just the RSS as a measure of quality-of-fit, and
then go back to the interesting cases to look at the full set of
coefficients and SEs.

Here, we present a few benchmarks comparing a couple of different
routines from the [Eigen library](http://eigen.tuxfamily.org)
(QR decomposition with column pivoting and Cholesky decomposition; the
latter doesn't handle the rank-deficient case). We compare these to
the [LAPACK](http://www.netlib.org/lapack/) routines used in
[R/qtl](http://rqtl.org):
[`dgels`](http://www.netlib.org/lapack/explore-html/d8/dde/dgels_8f.html)
and
[`dgelsy`](http://www.netlib.org/lapack/explore-html/d6/d4b/dgelsy_8f.html).
We also consider R's `lm.fit`, which is the workhorse behind `lm`.

These linear regression functions are not exported, so we need to use
`qtl2:::`.

## Simple case

First let's look at a simple case, of a single covariate.

```{r sim_simple_data}
set.seed(27343534)
n <- 1000
x <- rnorm(n, 50, 10)
X <- cbind(1, x)
y <- 30 + 0.5*x + rnorm(n, 0, 2.5)
Y <- as.matrix(y)
```

We use the
[microbenchmark package](http://cran.r-project.org/web/packages/microbenchmark/)
and run each of the routines 1000 times.

```{r simple_benchmark}
library(qtl2)
library(microbenchmark)
microbenchmark(lmfit =        lm.fit(X, y),
               dgels=         qtl2:::calc_rss_lapack(X, Y),
               dgelsy=        qtl2:::calc_rss_lapack(X, Y, skip_dgels=TRUE),
               eigen_qr_rss=  qtl2:::calc_rss_eigenqr(X, y),
               eigen_qr=      qtl2:::fit_linreg_eigenqr(X, y),
               eigen_chol_rss=qtl2:::calc_rss_eigenchol(X, y),
               eigen_chol=    qtl2:::fit_linreg_eigenchol(X, y),
               times=1000)
```

It's easy to beat `lm.fit`, but it's great to see considerable gains
with Eigen over LAPACK. The use of the Cholesky decomposition is
somewhat faster than QR decomposition, but it can't be used in the
rank-deficient case.

## Rank-deficient case

Let's now consider a rank-deficient case. We use an example from 
[Bates and Eddelbuettel (2013)](http://www.jstatsoft.org/v52/i05/)
(the [RcppEigen](https://github.com/RcppCore/RcppEigen) paper).

```{r sim_rankdef_data}
dd <- data.frame(f1 = gl(4, 6, labels = LETTERS[1:4]),
                 f2 = gl(3, 2, labels = letters[1:3]))[-(7:8), ]
mm <- model.matrix(~ f1*f2, dd)
y <- mm %*% seq_len(ncol(mm)) + rnorm(nrow(mm), sd = 0.1)
Y <- as.matrix(y)
```

And here are the benchmark results.

```{r rankdef_benchmark}
microbenchmark(lmfit=       lm.fit(mm, y),
               dgels=       qtl2:::calc_rss_lapack(mm, Y),
               dgelsy=      qtl2:::calc_rss_lapack(mm, Y, skip_dgels=TRUE),
               eigen_qr_rss=qtl2:::calc_rss_eigenqr(mm, y),
               eigen_qr=    qtl2:::fit_linreg_eigenqr(mm, y),
               times=1000)
```

We skip the methods based on Cholesky decomposition, but we again see
performance gains with Eigen over LAPACK.

`dgels` is slower than `dgelsy` here, because with that routine, I'm
calling `dgels` to check the matrix rank and then switch to `dgelsy`
if the matrix is rank-deficient.


## Bigger matrix

Let's consider a bigger matrix, with 2,500 rows and 300 columns. I
don't want it to be too big, or compiling this vignette will take too
long. Try this yourself with bigger matrices, if you want.

```{r sim_big}
n <- 2500
p <- 300
X <- matrix(rnorm(n*p), nrow=n)
y <- X %*% runif(p) + rnorm(n)
Y <- as.matrix(y)
```

We use the full set again, as we can assume this to be a full-rank
case. But we'll just do 10 replicates.

```{r big_benchmark}
microbenchmark(lmfit =        lm.fit(X, y),
               dgels=         qtl2:::calc_rss_lapack(X, Y),
               dgelsy=        qtl2:::calc_rss_lapack(X, Y, skip_dgels=TRUE),
               eigen_qr_rss=  qtl2:::calc_rss_eigenqr(X, y),
               eigen_qr=      qtl2:::fit_linreg_eigenqr(X, y),
               eigen_chol_rss=qtl2:::calc_rss_eigenchol(X, y),
               eigen_chol=    qtl2:::fit_linreg_eigenchol(X, y),
               times=10)
```

## Multiple phenotypes

We'll go back to the case with a single covariate, but we'll consider
500 phenotypes. We have a few routines to handle this, focusing
solely on calculating the residual sum of squares for each phenotype.

```{r sim_multi_pheno}
n <- 1000
x <- rnorm(n, 50, 10)
X <- cbind(1, x)
y <- 30 + 0.5*x + rnorm(n, 0, 2.5)
Y <- qtl2:::permute_nvector(500, y)
```

Now we'll get the benchmarks.

```{r benchmark_multi_pheno}
microbenchmark(lmfit =    colSums(lm.fit(X, Y)$resid^2),
               dgels=     qtl2:::calc_rss_lapack(X, Y),
               dgelsy=    qtl2:::calc_rss_lapack(X, Y, skip_dgels=TRUE),
               eigen_qr=  qtl2:::calc_mvrss_eigenqr(X, Y),
               eigen_chol=qtl2:::calc_mvrss_eigenchol(X, Y),
               times=50)
```


## Multiple phenotypes, rank-deficient matrix

Let's do the same thing with a rank-deficient matrix.

```{r multi_pheno_rank_def}
dd <- data.frame(f1 = gl(4, 6, labels = LETTERS[1:4]),
                 f2 = gl(3, 2, labels = letters[1:3]))[-(7:8), ]
mm <- model.matrix(~ f1*f2, dd)
y <- mm %*% seq_len(ncol(mm)) + rnorm(nrow(mm), sd = 0.1)
Y <- qtl2:::permute_nvector(500, y)
```

And here are the benchmarks.

```{r benchmark_multi_pheno_rank-def}
microbenchmark(lmfit =    colSums(lm.fit(mm, Y)$resid^2),
               dgels=     qtl2:::calc_rss_lapack(mm, Y),
               dgelsy=    qtl2:::calc_rss_lapack(mm, Y, skip_dgels=TRUE),
               eigen_qr=  qtl2:::calc_mvrss_eigenqr(mm, Y),
               times=100)
```

## Session information

The following shows the R and package versions I was using.

```{r sessionInfo, results="markup"}
devtools::session_info()
```

<!-- the following to make it look nicer -->
<link href="http://kbroman.org/qtlcharts/assets/vignettes/vignette.css" rel="stylesheet"></link>
